### 2022-09-19 chapter1/chapter.tex
finito

Following \cite{rossBook}, a stochastic process is a sequence, \(\{X(t)\}_{t\in\mathcal T}\), of random variables (or random vectors) indexed by some index set \(\mathcal T\). In this thesis, in the case that the index set is \(\mathcal T=\{0,1,2,...\}\) we say that the process \(\{X(n)\}_{n\in \mathcal T}\) is a discrete-time process, and we will typically use the dummy variable \(n\) for the `time' index. When the index set is \(\mathcal T=[0,\infty)\) we say \(\{X(t)\}_{t\in\mathcal T}\) is a continuous-time process, and we will typically use the dummy variable \(t\) for the `time' index. We may omit the index set and write \(\{X(t)\}\) in place of \(\{X(t)\}_{t\in\mathcal T}\) when it is not explicitly needed, or we may write \(\{X(t)\}_{t\geq t_0}\) to mean \(\{X(t)\}_{t\in[t_0,\infty)}\). The \emph{state space} of \(X(t)\) is the set of possible values that \(X(t)\) can take at any time \(t\in\mathcal T\).

@@

A stochastic process is \emph{stationary} if, for any \(n\), \(t_0<t_1<...<t_n\) and any \(t\), \(X(t_n),X(t_{n-1}),...,X(t_0)\) and \(X(t_n+t),X(t_{n-1}+t),...,X(t_0+t)\) have the same distribution. That is, 
\begin{align}\label{eqn: stationary}
	&\nonumber\mathbb P(X(t_n)\in E_n,X(t_{n-1})\in E_{n-1},...,X(t_0)\in E_0) 
	\\&= \mathbb P(X(t_n+t)\in E_n,X(t_{n-1}+t)\in E_{n-1},...,X(t_0+t)\in E_0),
\end{align}
for any \(t,t_0,...,t_n\in\mathcal T\) with \(t_0+t,...,t_n+t\in\mathcal T\) and any \(E_0,...,E_n\subseteq \mathcal S\). A random variable, \(\tau\), is a \emph{stopping time} for the stochastic process \(\{X(t)\}_{t\in\mathcal T}\) if \(\tau\in\sigma(X(s), s\leq t),\) where \(\sigma(X(s), s\leq t)\) denotes the \(\sigma\)-algebra generated by \(\{X(s), s\leq t\}\).

@@

Also, as in \cite[Section~1.3]{MEinAP}, we call \(\{X(t)\}_{t\geq 0}\) a \emph{continuous-time Markov chain} (CTMC) if \(\calS\) is countable and for all \(t_{n+1} > t_n > t_{n-1} > ... > t_0 > 0\) and \(i,i_0,...,i_{n-1},j\in\calS,\)
\begin{align}\label{eqn: Markov property cmtc}
	\mathbb P(X(t_{n+1})=j \mid X(n)=i, X(n-1)=i_{n-1},...,X(0)=i_0) = \mathbb P(X(t_{n+1})=j \mid X(t_n)=i),
\end{align}
and refer to (\ref{eqn: Markov property cmtc}) as the \emph{Markov property}. The process \(\{X(t)\}_{t\geq 0}\) is said to be time-homogeneous if 
\(\mathbb P(X(t+s)=j \mid X(s)=i)\) does not depend on \(s\). For a time-homogeneous CTMC, the \emph{transition function} is the matrix function 
\[\bs P(t) = [\mathbb P(X(t)=j \mid X(t)=i)]_{i,j\in\calS}.\]
The Chapman-Kolmogorov equations state that 
\begin{align} 
	&\nonumber \bs P(t+s)=\bs P(t)\bs P(s).
\end{align}
The \emph{infinitesimal generator} matrix of a CTMC is 
\begin{align} 
	&\nonumber \bs T = [T_{ij}]_{i,j\in\calS} = \cfrac{\wrt}{\wrt t}P(t),
\end{align}
and the elements \(T_{ij}\) are the \emph{transition rates}. A CTMC is said to be \emph{honest} if \(\bs P(t)\bs 1=\bs 1\) for all \(t\geq 0\) where \(\bs 1\) is a column vector of ones.

@@

Most generally, we call \(\{X(t)\}_{t\geq 0}\) a \emph{Markov process} if for all \(t_n > t_{n-1} > ... > t_0 > 0,\, t_k\in\mathcal T,\, k=0,...,n+1,\) and \(E_0,...,E_n\subseteq \mathcal S\)
\begin{align}\label{eqn: Markov process}
	\mathbb P(X(t_n)\in E_n,X(t_{n-1})\in E_{n-1},...,X(t_0)\in E_0) 
	= \mathbb P(X(t_n)\in E_n\mid X(t_{n-1})),
\end{align}
and is time homogeneous if \(\mathbb P(X(t+s)\in E_n\mid X(s))\) does not depend on \(s\).

@@

Let \(c\in \mathbb R\) and \(v\in V\), \(w\in W\). Then \(\mathbb B\) is a \emph{linear} operator if \(\mathbb B(cv) = c(\mathbb Bv)\) and \(\mathbb B(v+w)=\mathbb Bv+\mathbb Bw\). We define the norm of an operator as \(||\mathbb B||=\sup_{v\in V}\cfrac{||\mathbb Bv||_W}{||v||_V}\), where \(||\cdot||_V\) and \(||\cdot||_W\) are the norms defined on \(V\) and \(W\), respectively.  An operator is said to be bounded if \(||\mathbb B||<M\) for some \(M<\infty\).

@@

Consider the expectation
\[\mathbb E[f(X(t))\mid X(0)=x]=\int_{y\in\calS} f(y) P(X(t)\in\wrt y\mid X(0)=x),\]
where \(f<F\) is some bounded function. We can think of expectation as an operator which is acting on the function \(f\). The expectation depends on \(t\) and \(x\) as these define the distribution with which we are taking the expectation, hence, let us write 
\[\mathbb T(t)f(x) = \mathbb E[f(X(t))\mid X(0)=x].\]
The result of the operator acting on \(f\), \(\mathbb T(t)f(x)\), is another function.

@@

The level process is given by 
\[\ddot X(t) = \ddot X(0) + \int_{s=0}^t c_{\varphi(s)}\wrt s.\]
Sample paths of \(\{\ddot X (t)\}\) are continuous and piecewise linear, with \(\cfrac{\wrt }{\wrt t} \ddot X(t) = c_\varphi(t)\), when \(\ddot X(t)\) is differentiable. Given sample paths of \(\{\varphi(t)\}\), then \(\{\ddot X(t)\}\) is deterministic, and in this sense, \(\{\varphi(t)\}\) is the only stochastic element of the fluid queue.

@@

\subsection{Transient analysis of fluid queues}\label{sec: transient ffq intro}
The transient analysis of fluid queues has been relatively well studied (see, for example, \cite{ar2004,bean2005,dasilva2005,bean2009} which are matrix-analytics-methods-based, and also \cite{rs2003} which is differential-equations-based and treats a slightly more complex model than the ones considered here, but is none-the-less relevant). Given the interest in discontinuous and point-mass initial conditions, transient analysis of fluid queues rarely relies on solving governing differential equations numerically, although it is quite possible to do so\footnote{We have already alluded to some difficulties with differential-equation-based approaches to problems with discontinuous solutions such as ill-defined PDEs, the need for weak solutions, and oscillatory approximations with possibly infeasible approximate solutions, for example negative probability approximations.}. Instead, expressions for transient distributions of fluid queues are derived in terms of Laplace transforms with respect to time, and/or moments of the transient distributions are derived \citep{ar2004,bean2005,bean2009}. Moreover, these techniques often lead to quantities which have direct probabilistic interpretations which further aids in their analysis and also applications to other problems \citep{ar2003,dasilva2005,bnp2018} (we also leverage these stochastic interpretations in Chapter~\ref{sec: conv} to derive expressions for certain the Laplace transforms of the fluid queue). In some contexts, the actual transient distributions are not required, and Laplace transforms or moments are all that are required. In other cases, we can invert the Laplace transforms using known methods (\cite{aw2006}, or in the case of the discontinuities we might prefer \cite{hhat2020}, which uses the same concentrated matrix exponential distribution that we do).

@@

It turns out that, for fixed \(\lambda \geq 0\), \(\bs H (\lambda,y)\) is a semigroup (with variable \(y\)) \citep{bean2005}. \cite{bean2005} find the infinitesimal generator, \(\bs Q(\lambda)\) of \(\bs H(\lambda,y)\) and, since the generator is bounded, we can write \(\bs H(\lambda,y)=e^{\bs Q(\lambda)y}\). The derivation of the generator \(\bs Q(\lambda)=[Q_{ij}(\lambda)]_{ij\in\calS_+\cup\calS_-}\) is based on a direct analysis of sample paths of the fluid queue over an infinitesimal time, \(u\) say, and by leveraging the fact that complex sample paths occur with probability \(\mathcal O(u^2)\). As a result, there are only three types of sample paths to consider. Further, \cite{bean2005} similarly argue that the generator \(\bs Q(\lambda)\) can be partitioned into blocks \(\bs Q_{mn}(\lambda) = \left[Q_{ij}(\lambda)\right]_{i\in\calS_m,j\in\calS_n}\), \(m\in\{+,-\},\,n\in\{+,-,0\}\) where 
\begin{align*}
	\bs Q_{+0}(\lambda) &= \bs C_+^{-1}\bs T_{+0}\left[\lambda \bs I - \bs T_{00}\right]^{-1},
	%
	\\\bs Q_{-0}(\lambda) &= \bs C_-^{-1}\bs T_{-0}\left[\lambda \bs I - \bs T_{00}\right]^{-1},
	%
	\\\bs Q_{++}(\lambda) &= \bs C_+^{-1} \left(\bs T_{++} - \lambda \bs I + \bs T_{+0}\left[\lambda \bs I - \bs T_{00}\right]^{-1}\bs T_{0+}\right),
	%
	\\\bs Q_{+-}(\lambda) &= \bs C_+^{-1} \left(\bs T_{+-} + \bs T_{+0}\left[\lambda \bs I - \bs T_{00}\right]^{-1}\bs T_{0-} \right) ,
	%
	\\\bs Q_{--}(\lambda) &= \bs C_-^{-1} \left(\bs T_{--}  - \lambda \bs I + \bs T_{-0}\left[\lambda \bs I - \bs T_{00}\right]^{-1}\bs T_{0-}\right),
	%
	\\\bs Q_{-+}(\lambda) &=\bs C_-^{-1} \left(\bs T_{-+}+ \bs T_{-0}\left[\lambda \bs I - \bs T_{00}\right]^{-1}\bs T_{0+}\right).
\end{align*}
\cite{bean2005} also define the functions 
\begin{align}
	\bs H^{++}(\lambda,y)&= \left[h_{ij}^{++}(\lambda,y)\right]_{i\in \mathcal S_+,j\in\mathcal S_{+}\cup\calS_{+0}} := e^{\bs{Q}_{++}(\lambda)y}\vligne{\bs C_+^{-1} & \bs Q_{+0}(\lambda)},  \label{eqn: lst 1a}
	\\\bs H^{--}(\lambda,y) &= \left[h_{ij}^{--}(\lambda,y)\right]_{i\in\mathcal S_-,j\in\mathcal S_{-}\cup\calS_{-0}}:= e^{\bs{Q}_{--}(\lambda)y}\vligne{\bs C_-^{-1} & \bs Q_{-0}(\lambda)},
	\\\bs H^{+-}(\lambda,y)  &= \left[h_{ij}^{+-}(\lambda,y)\right]_{i\in\mathcal S_+,\,j\in\mathcal S_-}:= e^{\bs{Q}_{++}(\lambda)y}\bs{Q}_{+-}(\lambda), 
	\\\bs H^{-+}(\lambda,y)&= \left[h_{ij}^{-+}(\lambda,y)\right]_{i\in\mathcal S_-,\, j\in\mathcal S_+} := e^{\bs{Q}_{--}(\lambda)y}\bs{Q}_{-+}(\lambda), \label{eqn: lst 4a}
\end{align}
for \(y,\lambda\geq 0,\) which have the following stochastic interpretations. The function \(h_{ij}^{++}(\lambda,y)\) (\(h_{ij}^{--}(\lambda,y)\)) is the Laplace transform (with respect to time) of the time taken for the fluid level to shift by an amount \(y\) whilst remaining in phases in \(\mathcal S_+\cup\calS_{+0}\) (\(\mathcal S_-\cup\calS_{-0}\)), given the phase was initially \(i\in\mathcal S_+\) (\(i\in\mathcal S_-\)). The function \(h_{ij}^{+-}(\lambda,y)\) (\(h_{ij}^{-+}(\lambda,y)\)) is the Laplace transform (with respect to time) of the time taken for the fluid level, \(\{Y(t)\}\) to shift by an amount \(y\) whilst remaining in phases in \(\mathcal S_+\cup\calS_{+0}\) (\(\mathcal S_-\cup\calS_{-0}\)), after which time the phase instantaneously changes to \(j\in\mathcal S_-\) (\(\mathcal S_+\)), given the phase was initially \(i\in\mathcal S_+\) (\(\mathcal S_-\)) \citep{bean2005}.

Thus, \cite{bean2005} are able to characterise, by the above matrix expressions, segments of sample paths of the fluid queue where the fluid level is non-decreasing and non-increasing. \cite{bean2005} then partition the sample paths which contribute to \(\bs \Psi_X\) as those which either, (a) have a single transition from \(\calS_+\) to \(\calS_-\) (perhaps via \(\calS_0\)), and, (b) those which have more than one transition from \(\calS_+\) to \(\calS_-\) (perhaps via \(\calS_0\)). An expression for Laplace-Stieltjes transform of (a) is 
\[\int_{y=0}^\infty e^{\bs Q_{++}(\lambda)y}\bs Q_{+-}(\lambda)e^{\bs Q_{--}y}\wrt y.\]
For the sample paths (b), there must be at least one point at which the phase process transitions from \(\calS_-\) to \(\calS_+\) (perhaps via \(\calS_0\)) before the first return time. Further, one of the transitions from \(\calS_-\) to \(\calS_+\) (perhaps via \(\calS_0\)) must occur lower than all others. By considering the lowest level, \(y\), at which a phase transitions from \(\calS_-\) to \(\calS_+\) (perhaps via \(\calS_0\)) occurs, \cite{bean2005} characterise the Laplace transform of the paths (b) by the expression 
\begin{align}
	\int_{y=0}^\infty e^{\bs Q_{++}(\lambda)y} \bs \Psi_x(\lambda)\bs Q_{-+}(\lambda)\bs \Psi_X(\lambda) e^{\bs Q_{--}(\lambda)y}\wrt y.
\end{align}

@@

The concepts leading to the derivation of \( {\mathbb \Psi}\) for the fluid-fluid queue are much the same as the derivation of the analogous quantity, \(\bs\Psi_X\), for a classical fluid queue. First, we need the infinitesimal generator of the driving process, \(\{(X(t),\varphi(t))\}\), which we denote by \(\mathbb B\). Then we need to partition \(\mathbb B\) on sets for which \(\{\dot Y(t)\}\) is increasing, decreasing or constant -- these are the sets \(\mathcal F_i^m,\, i\in\calS,\,m\in\{+,-,0\}\), below. We can then derive an operator-Riccati equation for the first return operator for the fluid-fluid queue, \(\mathbb \Psi\), in terms of the partitioned generator \(\mathbb B\), the rates \(r_i(x)\). The solution to the operator-Riccati equation gives the distribution of the driving process \(\{(X(t),\varphi(t))\}\) at the time when \(\{\dot Y(t)\}\) first returns to \(0\).

The problem we have is to solve the operator-Riccati equation -- something which is only possibly in the very simplest of cases. This is where the methods in this thesis come in. Here, we approximate \(\mathbb B\) by a finite-dimensional matrix which we then partition according to the sets \(\mathcal F_i^m,\, i\in\calS,\,m\in\{+,-,0\}\). We then substitute the resulting matrices into the operator-Riccati equation and this gives us a matrix-Riccati equation which we can then solve using known methods \cite{bot08}.

@@

We assume that the process $\{(\dot Y(t), X(t), \varphi(t))\}_{t\geq 0}$ is positive recurrent, in order to guarantee the existence of the joint limiting distribution. Define limiting operators 
\begin{align} 
		\label{eqn:jointpi} 
		\bbpi_i(y)(\mathcal{E}) & := \lim_{t \rightarrow \infty} \frac{\partial}{\partial y} \mathbb{P}\left( \dot Y(t) \leq y,  X(t) \in \mathcal{E}, \varphi(t) = i\right), \quad y>0,\\
% 
		\label{eqn:jointmass}
		{\mathbb p}_i(\mathcal{E}) & := \lim_{t \rightarrow \infty}  \mathbb{P}( \dot Y(t) = 0,  X(t) \in \mathcal{E}, \varphi(t) = i]),
\end{align} 
where $\mathcal{E} \subset [0,b]$. The keen observer will notice that (\ref{eqn:jointpi}) assumes a certain differentiability condition on the limiting distribution. Indeed, it is not always the case that (\ref{eqn:jointpi}) will be differentiable with respect to \(y\) (trivially, consider \(\dot Y(t)=y\) for all \(t\)). Conditions on fluid-fluid queues to ensure that (\ref{eqn:jointpi}) will be differentiable with respect to \(y\) are not known, and here we just assume that (\ref{eqn:jointpi}) is differentiable.

@@

The generator of a fluid queue is a differential operator and to enable computation of the first-return operator, approximation methods are needed. The approximation schemes which we discuss in this thesis are all cell-based methods which discretise the level of the fluid queue into cells. Thus, one complexity in approximation is to reconcile the partition according to whether the second fluid is increasing, decreasing or constant and the partition which the approximation method uses. As long as the rates of the fluid-fluid queue are not `too-badly' behaved, this is not terribly difficult to do, but it does require some discussion and notation.

@@

For CMEs considered here, the vector function \(\bs k(t)=\cfrac{\bs \alpha e^{St}}{\bs \alpha e^{St}\bs e}\) is periodic with period \(\rho = 2\pi/\omega\) where \(\omega=min_i(|\Im(\lambda_i)|)\), \(\lambda_i\) are the eigenvalues of \(\bs{S}\) and \(\Im(z)\) is the imaginary component of a complex number \(z\). We leverage this fact to simplify some numerical integration in Chapter~\ref{sec: construction and modelling}.

>>>>

Following \cite{rossBook}, a stochastic process is a sequence, \(\{X(t)\}_{t\in\mathcal T}\), of random variables (or random vectors) indexed by some index set \(\mathcal T\). In this thesis, in the case that the index set is \(\mathcal T=\{0,1,2,...\}\) we say that the process \(\{X(n)\}_{n\in \mathcal T}\) is a discrete-time process, and we will typically use the dummy variable \(n\) for the `time' index. When the index set is \(\mathcal T=[0,\infty)\) we say \(\{X(t)\}_{t\in\mathcal T}\) is a continuous-time process, and we will typically use the dummy variable \(t\) for the `time' index. We may omit the index set and write \(\{X(t)\}\) in place of \(\{X(t)\}_{t\in\mathcal T}\) when it is not explicitly needed, or we may write \(\{X(t)\}_{t\geq t_0}\) to mean \(\{X(t)\}_{t\in[t_0,\infty)}\). The \emph{state space}, \(\mathcal S\), of \(X(t)\) is the set of possible values that \(X(t)\) can take at any time \(t\in\mathcal T\).

@@

A stochastic process is \emph{stationary} if, for any \(n\), \(t_0<t_1<...<t_n\) and any \(t\), then \(X(t_n),X(t_{n-1}),...,X(t_0)\) and \(X(t_n+t),X(t_{n-1}+t),...,X(t_0+t)\) have the same distribution. That is, 
\begin{align}\label{eqn: stationary}
	&\nonumber\mathbb P(X(t_n)\in E_n,X(t_{n-1})\in E_{n-1},...,X(t_0)\in E_0) 
	\\&= \mathbb P(X(t_n+t)\in E_n,X(t_{n-1}+t)\in E_{n-1},...,X(t_0+t)\in E_0),
\end{align}
for any \(t,t_0,...,t_n\in\mathcal T\) with \(t_0+t,...,t_n+t\in\mathcal T\) and any \(E_0,...,E_n\subseteq \mathcal S\). A random variable, \(\tau\), is a \emph{stopping time} for the stochastic process \(\{X(t)\}_{t\in\mathcal T}\) if \(\tau\in\sigma(X(s), s\leq t),\) where \(\sigma(X(s), s\leq t)\) denotes the \(\sigma\)-algebra generated by \(\{X(s), s\leq t\}\).

@@

Also, as in \cite[Section~1.3]{MEinAP}, we call \(\{X(t)\}_{t\geq 0}\) a \emph{continuous-time Markov chain} (CTMC) if \(\calS\) is countable and for all \(t_{n+1} > t_n > t_{n-1} > ... > t_0 \geq 0\) and \(i,i_0,...,i_{n-1},j\in\calS,\)
\begin{align}\label{eqn: Markov property cmtc}
	\mathbb P(X(t_{n+1})=j \mid X(t_n)=i, X(t_{n-1})=i_{n-1},...,X(t_0)=i_0) = \mathbb P(X(t_{n+1})=j \mid X(t_n)=i),
\end{align}
and refer to (\ref{eqn: Markov property cmtc}) as the \emph{Markov property}. The process \(\{X(t)\}_{t\geq 0}\) is said to be time-homogeneous if 
\(\mathbb P(X(t+s)=j \mid X(s)=i)\) does not depend on \(s\). For a time-homogeneous CTMC, the \emph{transition function} is the matrix function 
\[\bs P(t) = [\mathbb P(X(t)=j \mid X(t)=i)]_{i,j\in\calS}.\]
The Chapman-Kolmogorov equations state that 
\begin{align} 
	&\nonumber \bs P(t+s)=\bs P(t)\bs P(s).
\end{align}
The \emph{infinitesimal generator} matrix of a CTMC is 
\begin{align} 
	&\nonumber \bs T = [T_{ij}]_{i,j\in\calS} = \cfrac{\wrt}{\wrt t}P(t),
\end{align}
and the elements \(T_{ij}\) are the \emph{transition rates}. A CTMC is said to be \emph{honest} if \(\bs P(t)\bs 1=\bs 1\) for all \(t\geq 0\) where \(\bs 1\) is a column vector of ones.

@@

Most generally, we call \(\{X(t)\}_{t\geq 0}\) a \emph{Markov process} if for all \(t_n > t_{n-1} > ... > t_0 \geq 0,\, t_k\in\mathcal T,\, k=0,...,n,\) and \(E_0,...,E_n\subseteq \mathcal S\)
\begin{align}\label{eqn: Markov process}
	\mathbb P(X(t_n)\in E_n,\mid X(t_{n-1})\in E_{n-1},...,X(t_0)\in E_0) 
	= \mathbb P(X(t_n)\in E_n\mid X(t_{n-1})),
\end{align}
and we say that it is time homogeneous if \(\mathbb P(X(t+s)\in E_n\mid X(s))\) does not depend on \(s\).

@@

Let \(c\in \mathbb R\) (\(\mathbb R\) is the reals), \(v\in V\) and \(w\in W\). Then \(\mathbb B\) is a \emph{linear} operator if \(\mathbb B(cv) = c(\mathbb Bv)\) and \(\mathbb B(v+w)=\mathbb Bv+\mathbb Bw\). We define the norm of an operator as \(||\mathbb B||=\sup_{v\in V}\cfrac{||\mathbb Bv||_W}{||v||_V}\), where \(||\cdot||_V\) and \(||\cdot||_W\) are the norms defined on \(V\) and \(W\), respectively.  An operator is said to be bounded if \(||\mathbb B||<M\) for some \(M<\infty\).

@@

Consider the expectation
\[\mathbb E[f(X(t))\mid X(0)=x]=\int_{y\in\calS} f(y) P(X(t)\in\wrt y\mid X(0)=x),\]
where \(f\) is some bounded function, \(|f|<F\). We can think of expectation as an operator which is acting on the function \(f\). The expectation depends on \(t\) and \(x\) as these define the distribution with which we are taking the expectation, hence, let us write 
\[\mathbb T(t)f(x) = \mathbb E[f(X(t))\mid X(0)=x].\]
The result of the operator acting on \(f\), \(\mathbb T(t)f\), is another function (a function of the initial point \(x\)).

@@

Now, back to fluid queues. The level process is given by 
\[\ddot X(t) = \ddot X(0) + \int_{s=0}^t c_{\varphi(s)}\wrt s.\]
Sample paths of \(\{\ddot X (t)\}\) are continuous and piecewise linear, with \(\cfrac{\wrt }{\wrt t} \ddot X(t) = c_\varphi(t)\), when \(\ddot X(t)\) is differentiable. Given sample paths of \(\{\varphi(t)\}\), then \(\{\ddot X(t)\}\) is deterministic, and in this sense, \(\{\varphi(t)\}\) is the only stochastic element of the fluid queue.

@@

\subsection{Transient analysis of fluid queues}\label{sec: transient ffq intro}
The transient analysis of fluid queues has been relatively well studied (see, for example, \cite{ar2004,bean2005,dasilva2005,bean2009} which are matrix-analytics-methods-based, and also \cite{rs2003} which is differential-equations-based and treats a slightly more complex model than the ones considered here, but is none-the-less relevant). Given the interest in discontinuous and point-mass initial conditions, transient analysis of fluid queues rarely relies on solving governing differential equations numerically, although it is quite possible to do so\footnote{We have already alluded to some difficulties with differential-equation-based approaches to problems with discontinuous solutions such as ill-defined PDEs, the need for weak solutions, and oscillatory approximations with possibly infeasible approximate solutions, for example negative probability approximations.}. Instead, expressions for transient distributions of fluid queues are derived in terms of Laplace transforms with respect to time, and/or moments of the transient distributions are derived \citep{ar2004,bean2005,bean2009}. Moreover, these techniques often lead to quantities which have direct probabilistic interpretations which further aids in their analysis and also applications to other problems \citep{ar2003,dasilva2005,bnp2018} (we also leverage these stochastic interpretations in Chapter~\ref{sec: conv} to derive expressions for certain the Laplace transforms of the fluid queue). In some contexts, the actual transient distributions are not required, and Laplace transforms or moments, which can be obtained relatively straightforwardly from the Laplace transforms, are all that are required. In other cases, the Laplace transforms may be inverted using known methods (\cite{aw2006}, or in the case of the discontinuities we might prefer \cite{hhat2020}, which uses the same concentrated matrix exponential distribution that we do).

@@

It turns out that, for fixed \(\lambda \geq 0\), \(\bs H (\lambda,y)\) is a semigroup (with variable \(y\)) \citep{bean2005}. \cite{bean2005} find the infinitesimal generator, \(\bs Q(\lambda)\) of \(\bs H(\lambda,y)\) and, since the generator is bounded, we can write \(\bs H(\lambda,y)=e^{\bs Q(\lambda)y}\). The derivation of the generator \(\bs Q(\lambda)=[Q_{ij}(\lambda)]_{ij\in\calS_+\cup\calS_-}\) is based on a direct analysis of sample paths of the fluid queue over an infinitesimal time, \(u\) say, and by leveraging the fact that complex sample paths occur with probability \(\mathcal O(u^2)\). As a result, there are only three types of sample paths to consider. Further, \cite{bean2005} argue that the generator \(\bs Q(\lambda)\) can be partitioned into blocks \(\bs Q_{mn}(\lambda) = \left[Q_{ij}(\lambda)\right]_{i\in\calS_m,j\in\calS_n}\), \(m\in\{+,-\},\,n\in\{+,-,0\}\) where 
\begin{align*}
	\bs Q_{+0}(\lambda) &= \bs C_+^{-1}\bs T_{+0}\left[\lambda \bs I - \bs T_{00}\right]^{-1},
	%
	\\\bs Q_{-0}(\lambda) &= \bs C_-^{-1}\bs T_{-0}\left[\lambda \bs I - \bs T_{00}\right]^{-1},
	%
	\\\bs Q_{++}(\lambda) &= \bs C_+^{-1} \left(\bs T_{++} - \lambda \bs I + \bs T_{+0}\left[\lambda \bs I - \bs T_{00}\right]^{-1}\bs T_{0+}\right),
	%
	\\\bs Q_{+-}(\lambda) &= \bs C_+^{-1} \left(\bs T_{+-} + \bs T_{+0}\left[\lambda \bs I - \bs T_{00}\right]^{-1}\bs T_{0-} \right) ,
	%
	\\\bs Q_{--}(\lambda) &= \bs C_-^{-1} \left(\bs T_{--}  - \lambda \bs I + \bs T_{-0}\left[\lambda \bs I - \bs T_{00}\right]^{-1}\bs T_{0-}\right),
	%
	\\\bs Q_{-+}(\lambda) &=\bs C_-^{-1} \left(\bs T_{-+}+ \bs T_{-0}\left[\lambda \bs I - \bs T_{00}\right]^{-1}\bs T_{0+}\right).
\end{align*}
\cite{bean2005} also define the functions 
\begin{align}
	\bs H^{++}(\lambda,y)&= \left[h_{ij}^{++}(\lambda,y)\right]_{i\in \mathcal S_+,j\in\mathcal S_{+}\cup\calS_{+0}} := e^{\bs{Q}_{++}(\lambda)y}\vligne{\bs C_+^{-1} & \bs Q_{+0}(\lambda)},  \label{eqn: lst 1a}
	\\\bs H^{--}(\lambda,y) &= \left[h_{ij}^{--}(\lambda,y)\right]_{i\in\mathcal S_-,j\in\mathcal S_{-}\cup\calS_{-0}}:= e^{\bs{Q}_{--}(\lambda)y}\vligne{\bs C_-^{-1} & \bs Q_{-0}(\lambda)},
	\\\bs H^{+-}(\lambda,y)  &= \left[h_{ij}^{+-}(\lambda,y)\right]_{i\in\mathcal S_+,\,j\in\mathcal S_-}:= e^{\bs{Q}_{++}(\lambda)y}\bs{Q}_{+-}(\lambda), 
	\\\bs H^{-+}(\lambda,y)&= \left[h_{ij}^{-+}(\lambda,y)\right]_{i\in\mathcal S_-,\, j\in\mathcal S_+} := e^{\bs{Q}_{--}(\lambda)y}\bs{Q}_{-+}(\lambda), \label{eqn: lst 4a}
\end{align}
for \(y,\lambda\geq 0,\) which have the following stochastic interpretations. The function \(h_{ij}^{++}(\lambda,y)\) (\(h_{ij}^{--}(\lambda,y)\)) is the Laplace transform (with respect to time) of the time taken for the fluid level to shift by an amount \(y\) whilst remaining in phases in \(\mathcal S_+\cup\calS_{+0}\) (\(\mathcal S_-\cup\calS_{-0}\)), given the phase was initially \(i\in\mathcal S_+\) (\(i\in\mathcal S_-\)). The function \(h_{ij}^{+-}(\lambda,y)\) (\(h_{ij}^{-+}(\lambda,y)\)) is the Laplace transform (with respect to time) of the time taken for the fluid level, \(\{Y(t)\}\) to shift by an amount \(y\) whilst remaining in phases in \(\mathcal S_+\cup\calS_{+0}\) (\(\mathcal S_-\cup\calS_{-0}\)), after which time the phase instantaneously changes to \(j\in\mathcal S_-\) (\(\mathcal S_+\)), given the phase was initially \(i\in\mathcal S_+\) (\(\mathcal S_-\)) \citep{bean2005}.

Thus, \cite{bean2005} are able to characterise, by the above matrix expressions, segments of sample paths of the fluid queue where the fluid level is non-decreasing and non-increasing. \cite{bean2005} then partition the sample paths which contribute to \(\bs \Psi_X\) as those which either, (a) have a single transition from \(\calS_+\) to \(\calS_-\) (perhaps via \(\calS_0\)), and, (b) those which have more than one transition from \(\calS_+\) to \(\calS_-\) (perhaps via \(\calS_0\)). An expression for Laplace-Stieltjes transform of (a) is 
\[\int_{y=0}^\infty e^{\bs Q_{++}(\lambda)y}\bs Q_{+-}(\lambda)e^{\bs Q_{--}y}\wrt y.\]
For the sample paths (b), there must be at least one point at which the phase process transitions from \(\calS_-\) to \(\calS_+\) (perhaps via \(\calS_0\)) before the first return time. Further, one of the transitions from \(\calS_-\) to \(\calS_+\) (perhaps via \(\calS_0\)) must occur lower than all others. By considering the lowest level, \(y\), at which a phase transitions from \(\calS_-\) to \(\calS_+\) (perhaps via \(\calS_0\)) occurs, \cite{bean2005} characterise the Laplace transform of the paths (b) by the expression 
\begin{align}
	\int_{y=0}^\infty e^{\bs Q_{++}(\lambda)y} \bs \Psi_X(\lambda)\bs Q_{-+}(\lambda)\bs \Psi_X(\lambda) e^{\bs Q_{--}(\lambda)y}\wrt y.
\end{align}

@@

The concepts leading to the derivation of \( {\mathbb \Psi}\) for the fluid-fluid queue are much the same as the derivation of the analogous quantity, \(\bs\Psi_X\), for a classical fluid queue. First, we need the infinitesimal generator of the driving process, \(\{(X(t),\varphi(t))\}\), which we denote by \(\mathbb B\). Then we need to partition \(\mathbb B\) on sets for which \(\{\dot Y(t)\}\) is increasing, decreasing or constant -- these are the sets \(\mathcal F_i^m,\, i\in\calS,\,m\in\{+,-,0\}\), below. We can then derive an operator-Riccati equation for the first return operator for the fluid-fluid queue, \(\mathbb \Psi\), in terms of the partitioned generator and the rates \(r_i(x)\). The solution to the operator-Riccati equation is an operator which can be used to obtain the distribution of the driving process \(\{(X(t),\varphi(t))\}\) at the time when \(\{\dot Y(t)\}\) first returns to \(0\).

The problem we have is to solve the operator-Riccati equation -- something which is only possibly in the very simplest of cases. This is where the methods in this thesis come in. Here, we approximate \(\mathbb B\) by a finite-dimensional matrix which we then partition according to the sets \(\mathcal F_i^m,\, i\in\calS,\,m\in\{+,-,0\}\). We then substitute the resulting matrices into the operator-Riccati equation and this gives us a matrix-Riccati equation which we can then solve using known methods \citep{bot08}.

@@

We assume that the process $\{(\dot Y(t), X(t), \varphi(t))\}_{t\geq 0}$ is positive recurrent, in order to guarantee the existence of the joint limiting distribution. Define limiting operators 
\begin{align} 
		\label{eqn:jointpi} 
		\bbpi_i(y)(\mathcal{E}) & := \lim_{t \rightarrow \infty} \frac{\partial}{\partial y} \mathbb{P}\left( \dot Y(t) \leq y,  X(t) \in \mathcal{E}, \varphi(t) = i\right), \quad y>0,\\
% 
		\label{eqn:jointmass}
		{\mathbb p}_i(\mathcal{E}) & := \lim_{t \rightarrow \infty}  \mathbb{P}( \dot Y(t) = 0,  X(t) \in \mathcal{E}, \varphi(t) = i]),
\end{align} 
where $\mathcal{E} \subset [0,b]$. The keen observer will notice that (\ref{eqn:jointpi}) assumes a certain differentiability condition on the limiting distribution. Indeed, it is not always the case that (\ref{eqn:jointpi}) will be differentiable with respect to \(y\) (trivially, consider \(\dot Y(t)=y_0\) for all \(t\)). Conditions on fluid-fluid queues to ensure that (\ref{eqn:jointpi}) will be differentiable with respect to \(y\) are not known, and here we just assume that it is.

@@

The generator of a fluid queue is a differential operator and to enable computation of the first-return operator, approximation methods are needed. The approximation schemes which we discuss in this thesis are all cell-based methods which discretise the level of the fluid queue into intervals called cells. Thus, one complexity in approximation is to reconcile the partition according to whether the second fluid is increasing, decreasing or constant, and the partition which the approximation method uses. %As long as the rates of the fluid-fluid queue are not `too-badly' behaved, this is not terribly difficult to do, but it does require some discussion and notation.

@@

For CMEs the vector function \(\bs k(t)=\cfrac{\bs \alpha e^{St}}{\bs \alpha e^{St}\bs e}\) is periodic with period \(\rho = 2\pi/\omega\) where \(\omega=min_i(|\Im(\lambda_i)|)\), \(\lambda_i\) are the eigenvalues of \(\bs{S}\) and \(\Im(z)\) is the imaginary component of a complex number \(z\). We leverage this fact to simplify some numerical integration in Chapter~\ref{sec: construction and modelling}.
